---
format: typst
---

# Lab 4 Worksheet - Practice

```{r}
library(tidyverse)
library(dplyr)
library(stringr)
```

-   add something about loading a csv and having to use \# Split the one column into four energy_raw \<- separate( energy_raw, col = 1, into = c("State", "2021", "2022", "2023"), sep = "," )
-   make sure str extract is clear str_extract(.x, "\\d+\\.?\\d\*")

#### Q1. Extract and clean text patterns from strings using regular expressions.

You are given a messy dataset of energy sources with inconsistent units and spacing.

```{r}
energy_sources <- c("Solar (MWh)", "Wind (GWh)", "Coal (MWh)", "Hydro (kWh)")
```

a.  Use str_detect() to identify all energy sources that contain the substring "Wind" or "Hydro" (hint: use the \| operator in regex).

```{r}
str_detect(energy_sources, "Wind|Hydro")
```

b.  Extract the unit type (e.g., MWh, GWh, kWh) using str_extract(). Make sure your pattern handles extra spaces or hyphens.

```{r}
str_extract(energy_sources, "\\b[MGk]Wh\\b")
```

c.  Clean the names to keep only the energy source name, trimming whitespace and removing parentheses.

```{r}
sources_clean = str_replace(energy_sources, "\\s\\(.*\\)", "") |> str_trim()
sources_clean
```

d.  Create a clean tibble called energy_clean with two columns: source and unit.

```{r}
energy_clean = tibble(
    source=sources_clean,
    unit=str_extract(energy_sources, "[MGk]Wh")
)
energy_clean
```

e.  Discussion: Why might regex cleaning be important before performing joins or aggregations?\
    \
    Remove extraneous variables. Yay. Makes sure joins and analysis return correctly.

#### Q2. State Datasets: Joins and Pivots

We’ll use built-in datasets to explore US state-level data. state.x77 contains metrics such as population, income, and life expectancy. state.region classifies each state into one of four regions: Northeast, South, North Central, West.

```{r}
# Load datasets
library(dplyr)

# Convert matrix to data frame
states <- data.frame(state.x77)

# Add state names as a new column
states$state <- rownames(states)

# select and reorder columns
states <- states |>
  select(state, Income, Population, Illiteracy)
# table for regions
regions <- data.frame(state = state.name, region = state.region)

states
regions
```

a.  Perform a left join to combine socioeconomic and regional data.

```{r}
states_left_join <- states |> left_join(regions, by="state")
```

b.  Perform an inner join. How many rows remain, and why is it the same or different?

```{r}
states_inner_join <- states |> inner_join(regions, by="state")
#It's the same because both states and region have same states.
```

c.  Perform an anti join to check if any state names don’t match between datasets. What might cause mismatches like this in real projects?

```{r}
states_anti_join <- states |> anti_join(regions, by="state")
#0 mismatches, mismatches might come from typos, missing rows, etc. 
```

d.  Discussion: Why would you prefer a left join over an inner join when merging state-level datasets?\
    Left join -\> includes common data + all data from original data\
    Inner join -\> drop with missing matches, can bias
e.  Use pivot_longer() to make the socioeconomic data longer.

```{r}
states_longer <- states |> pivot_longer(
cols = c(Income, Population, Illiteracy),
names_to = "metric",
values_to = "value"
)
```

#### Q2.5.

Load this data:

```{r}
library(tidyverse)

# Sample sales data
sales <- tibble(
  Store = c("Store A", "Store B", "Store C"),
  Jan_Sales = c(100, 150, 120),
  Feb_Sales = c(200, 180, 130),
  Mar_Sales = c(150, 200, 140)
)

sales
```

a.  Convert the dataset from wide (one column per month) to long (one row per month per store).

```{r}
sales_long <- sales |> pivot_longer(cols=Jan_Sales:Mar_Sales, names_to="Month", values_to="Sales")
sales_long
```

b.  Clean month names by removing \_Sales.

```{r}
sales_long <- sales_long |>
    mutate(Month=str_remove(Month,"_Sales"))
sales_long
```

c.  Convert the long dataset back to wide, with one column per month.

```{r}
sales_wide <- sales_long |>
    pivot_wider(names_from=Month, values_from=Sales)
sales_wide
```

#### Q3. Mapping

Let's visualize data by state.

```{r}
library(ggplot2)
library(maps)
library(dplyr)

us_map <- map_data("state")

state_data <- states |>
  left_join(regions, by = "state") |>
  mutate(state = tolower(state))

us_map
state_data
```

Use a join to combine the us_map with the state_data

```{r}
us_map <- us_map |>
    rename(state = region)
us_map_joined <- left_join(state_data, us_map, by="state")
us_map_joined
```

Create a chloropleth map:

```{r}
ggplot(us_map_joined, aes(long, lat, group = group, fill = Income)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  theme_minimal() +
  scale_fill_viridis_c() +
  labs(title = "Average Income by State")
```

b.  Swap Income for Illiteracy to visualize illiteracy instead. Try changing the color scale using scale_fill_gradient() or scale_fill_gradient2()

```{r}
ggplot(us_map_joined, aes(long, lat, group = group, fill = Illiteracy)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  theme_minimal() +
  scale_fill_viridis_c() +
  labs(title = "Average Illiteracy by State")
```

#### Q4. Quarto Dashboard Setup

Create a new file ev-dashboard.qmd and add:

```{markdown}

---
title: "EV Power Dashboard"
format: dashboard
---

Add some sections:

### Overview

Your project introduction.

### Map Visualization

# Paste your ggplot map code here

### Data Table

# Show your joined dataset

head(us_map_joined)

Then render:

quarto preview

```